A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-07 podc_paper/algorithm-3.tex
working on GC

Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
We first describe the modifications to the implementation.
It uses two additional shared arrays.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ indexes last block in $root$ that process $k$ has dequeued from}
\end{itemize}
The \fld{blocks} field is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}},\\
		\gc{\fld{index} \assign\ \var{leaf}.\head}}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1,\\
	    \gc{\fld{index}\assign \var{leaf}.\head}}
    \State \Call{Append}{\var{B}}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}
    \label{deqRestGC}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}.\fld{blocks}.\Call{TryInsert}{\var{B}}\label{appendLeafGC}}
    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEndGC} 
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
    \State \var{h} \assign\ \var{v}.\head \label{readHeadGC}

	\gc{\If{ \var{v} is root and $\var{h} \mbox{ mod } p^2=0$}}
	\gc{\State \var{max} \assign 0}
	\gc{\For {$k \assign 1..p$}}
		\gc{\State $\var{max} \assign \max(\var{max}, \var{last}[k])$}
	\gc{\EndFor}
	\gc{\State \Call{Help}{}}
	\gc{\State \Call{Discard}{$\var{root}, \var{max}-1$}}
	\gc{\EndIf}

    \ForEach{\fld{dir} {\keywordfont{in}} \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHeadGC}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNullGC}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvanceGC}
        \EndIf
    \EndFor \label{endHelpChild1GC}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC} 
    \EndIf
    \gc{\State \var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryInsert}{\var{new}} \label{casGC}}
    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
    \State \Return{ \var{result}}
\EndFunction{Refresh}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

@@

\gc{
\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has been propagated to \var{root}}
    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]\neq \nl$}
    \If {$\var{v} = \var{root}$} \Return{true}
    \Else
    	\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
	    \State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
	    \If{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\edir$} \var{sup} \assign\ $\var{sup}+1$
	    \EndIf

	    \State \Return{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} \leq \var{v}.\fld{parent.blocks}[\var{sup}].\edir$ and \Call{Propagated}{\var{v}.\fld{parent}, \var{sup}}}
    \EndIf
\EndFunction{Propagated}

@@

\Function{void}{Help}{}
    \State \linecomment{help pending dequeues that have been propagated to root}
    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{h} \assign \ell.\head$
        \If{$\ell.\fld{blocks}[\var{h}].\fld{element} = \nl$ and \Call{Propagated}{$\ell, h$}}
            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, h, 1}$
            \State \var{output} \assign\ \Call{FindResponse}{$b, i$} 
            \State $\ell.\fld{blocks}[\var{h}].\fld{response} \assign\ \var{output}$
        \EndIf
    \EndFor
\EndFunction{Help}

\Function{void}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
    \State \linecomment{discard blocks with index less than or equal to threshold $t$ from RBT \var{v}.\fld{blocks} and all subblocks of those blocks}
    \State \var{t\sub{left}} \assign \var{v}.\fld{blocks}[\var{t}].\eleft
    \State \var{t\sub{right}} \assign \var{v}.\fld{blocks}[\var{t}].\eright
	\For{\var{j} \assign 1 to $p$}
        \State $T \assign \var{v}.\fld{blocks}$
        \State $T'  \assign \Call{Split}{T,t}$ \Comment{$T'$ is the root of new RBT containing $T$'s blocks with $\var{index} \geq t$}
        \State \bf{exit when} \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{splitCASGC}
    \EndFor
    \If{\var{v} is an internal \typ{Node}} \Comment{discard subblocks in descendants of $v$}
        \State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}
        \State \Call{Discard}{\var{v}.\fld{right}, \var{t\sub{right}}}
    \EndIf
\EndFunction{Discard}
}
\end{algorithmic}
\caption{\label{pseudocode2GC}Bounded-space queue implementation, continued. }
\end{figure}

@@

>>>>

Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
We first describe the modifications to the implementation.
It uses two additional shared arrays.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ indexes last block in $root$ that process $k$ has dequeued from}
\end{itemize}
The \fld{blocks} field is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}
We also apply an optimization:  instead of storing a \head\ field separate from \fld{blocks},
we can simply use the maximum \fld{index} of any block in the \fld{blocks} RBT.

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,$\\
		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}},$\\
		$\fld{index} \assign\ \var{leaf}.\head$}}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},$\\
	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1,$\\
	    $\fld{index}\assign \var{leaf}.\head$}}
    \State \Call{Append}{\var{B}}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head-1, 1}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}
    \label{deqRestGC}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{Insert}{\var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEndGC} 
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
    \gc{\State \here{Need to clean out my own leaf sometimes--how to determine what index to use?}}
\EndFunction{Append}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{\gc{GCRefresh}}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
        \State \Call{\gc{GCRefresh}}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\gc{\Function{RBT}{RefreshUpdate}{\typ{RBT} \var{T}, \typ{Node} \var{v}, \typ{int} \var{h}}}
    \State \linecomment{append new block to RBT $T$ read from $v.\fld{blocks}$}
    \ForEach{\fld{dir} {\keywordfont{in}} \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHeadGC}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNullGC}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvanceGC}
        \EndIf
    \EndFor \label{endHelpChild1GC}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \gc{\If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
    \Else \ \Return \Call{Insert}{$T$, \var{new}}
    \EndIf}
\EndFunction{RefreshUpdate}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{48}
\Function{boolean}{GCRefresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}, with GC}
    \State \var{h} \assign\ \var{v}.\head \label{readHeadGC}
	\gc{
	\State $\var{T} \assign\ \var{v}.\fld{blocks}$
	\If{max \fld{index} in $T$ is not $h-1$} 
		\State \linecomment \var{v}.\fld{blocks}[h] has already been added
		\State\Return{false}
	\ElsIf{\var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
		\State \linecomment{garbage collection phase}
		\State \var{max} \assign 0
		\For {$k \assign 1..p$}
			\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
		\EndFor
		\State \Call{Help}{}
		\State $\var{result} \assign \Call{Discard}{\var{root}, \var{max}-1}$
    \Else\ \linecomment{do a standard refresh}
        \State $T' \assign \Call{RefreshUpdate}{T, v, h}$
        \If{$T' = \nl$} \Return{\tr}
    	\Else\ $\var{result} \assign \Call{CAS}{\var{v}.\fld{blocks}, T, T'}$
		\EndIf
	\EndIf
	}
    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
    \State \Return{ \var{result}}
\EndFunction{Refresh}

@@

\gc{
%\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
%    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has been propagated to \var{root}}
%    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]\neq \nl$}
%    \If {$\var{v} = \var{root}$} \Return{true}
%    \Else
%    	\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
%	    \State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
%	    \If{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\edir$} \var{sup} \assign\ $\var{sup}+1$
%	    \EndIf
%	    \State \Return{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} \leq \var{v}.\fld{parent.blocks}[\var{sup}].\edir$ and \Call{Propagated}{\var{v}.\fld{parent}, \var{sup}}}
%    \EndIf
%\EndFunction{Propagated}
%
%\spac

\Function{void}{Help}{}
    \State \linecomment{help pending operations}

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{h} \assign \ell.\head$
        \If {$\ell.\fld{blocks}[h] = \nl$} $h \assign h-1$ \Comment{index of last block written in $\ell$}
        \EndIf
    	\State \Call{Propagate}{$\ell.\fld{parent}$}
        \If{$h>1$ and $\ell.\fld{blocks}[\var{h}].\fld{element} = \nl$} \Comment{operation is a  dequeue}
            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, h, 1}$
            \State $\ell.\fld{blocks}[\var{h}].\fld{response} \assign\ \Call{FindResponse}{b, i}$ 
        \EndIf
    \EndFor
\EndFunction{Help}

@@

\Function{boolean}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
    \State \linecomment{discard blocks with index less than or equal to threshold $t$ from RBT \var{v}.\fld{blocks} and all subblocks of those blocks}
    \If{$v$ is an internal node}
	    \State \var{t\sub{left}} \assign \var{v}.\fld{blocks}[\var{t}].\eleft
    	\State \var{t\sub{right}} \assign \var{v}.\fld{blocks}[\var{t}].\eright
		\For{\var{j} \assign 1 to $p$}
			\State $h \assign \var{v}.\head$
        	\State $T \assign \var{v}.\fld{blocks}$
        	\State $T'  \assign \Call{Split}{T,t}$ \Comment{$T'$ is the root of new RBT containing $T$'s blocks with $\var{index} \geq t$}
			\State $T'' \assign \Call{RefreshUpdate}{T',v,h}$ \Comment{do a \op{Refresh} too}
        	\State $\var{result} \assign \Call{CAS}{\var{v}.\fld{blocks}, T, (T'' = \nl\ ?\ T' : T'')$}\label{splitCASGC}
        	\State \bf{exit when} \var{result} \Comment{if \op{CAS} succeeds go to line \ref{Discardrecurse}}
    	\EndFor
    	\If{\var{v} is an internal \typ{Node}} \Comment{discard subblocks in descendants of $v$}\label{Discardrecurse}
        	\State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}
        	\State \Call{Discard}{\var{v}.\fld{right}, \var{t\sub{right}}}
    	\EndIf
	\EndIf
	\State \Return \var{result}
\EndFunction{Discard}
}
\end{algorithmic}
\caption{\label{pseudocode2GC}Bounded-space queue implementation, continued. }
\end{figure}

@@

\here{Remark on an optimization:  instead of storing \head\ field separate from \fld{blocks}, we could store
the maximum index of the RBT in the root of the RBT.  This would simplify things, but require modifying the proof more extensively.}
